Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.Rmd
some housekeeping and reordering

  The Rocker project provides widely-used Docker images for R across different application scenarios.

@@

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization basics and the Rocker Project, followed by descriptions of projects developing images with R installations.
Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This containerization is also called operating-system-level virtualization \ref{wikipedia_contributors_os-level_2020}.
Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
Nevertheless, special use cases and experimental projects exist to test approaches or provide features different to what _"base R"_ provides.
These projects stem both form academia and industry.
_Base R_, sometimes called `GNU-R`, is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and and provided via [CRAN](https://cran.r-project.org/).

@@

While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
All different distributions are published under GPL licenses.
Since all of the different R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

@@

In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker project provides containers pre-installed with the RStudio IDE.
In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.

@@

Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

@@

>>>>

  The Rocker&nbsp;Project provides widely-used Docker images for R across different application scenarios.

@@

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This containerization is also called operating-system-level virtualization \citep{wikipedia_contributors_os-level_2020}.
Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches or provide features different to what _GNU-R_, or Base R, provides.
GNU-R is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/).
While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
All these distributions of R are published under GPL licenses.
Since all of the R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.

@@

@@

In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker&nbsp;Project provides containers pre-installed with the RStudio IDE.
In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker&nbsp;Project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.

@@

Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker&nbsp;Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

@@

```{r word_stats, eval=FALSE, include=FALSE}
# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
wordcountaddin::readability(here::here("manuscript.Rmd"))
wordcountaddin::text_stats(here::here("manuscript.Rmd"))
```
